A geometric relationship of $F_2, F_3$ and $F_4$-statistics with Principal Component Analysis

### 2022-02-09 fstats_pca_lewontin.tex

The two data sets I analyzed here suggest that two PCs (for the World data set) respectively three PCs (for the Western Eurasian data) already provide a very good approximation for $F_4$-statistics (Figures \ref{fig:weu}E, \ref{fig:world}E), reflecting the observation that frequently the first few PCs provide a good approximation of population structure. On the other hand, for admixture $F_3$-statistics, more PCs are needed (Figure \ref{fig:weu}BC), likely because these statistics do contain a term measuring local variation in the putative admixed population \citep{peter2016}, which is typically often incorporated on higher PCs \citep{cavalli-sforza1975}. Thus, I would expect that approximating $F$-statistics with few PCs works well whenever local variation is unimportant, i.e. when calculating $F$-statistics far apart in PCA space. On the flip side, populations close-by on a PCA-plot are often seperated on higher PCs, and thus whenever close-by populations are included in the same $F$-statistics, these higher PCs will be required for good approximations.

@@

If sufficient PCs are considered, both $F_3$ and $F_4$-statistics do have simple interpretations in PCA-space. If used as a test for admixture, $F_3$ corresponds to a $n$-sphere in PCA-space, which manifests as a circle on a PCA-plot. The informal notion that an admixed population should lie between its sources on a PCA-plot is necessary, but not sufficient; unadmixed populations may also project there. In contrast, it is sufficient (but not necessary) to determine that a population has a positive $F_3$ statistic.

Interpretations of the outgroup-$F_3$-statistics and $F_4$-statistics in a PCA-framework rely heavily on the idea of a projection: In a tree or admixture-graph framework, we can loosely interpret $F_4(A, B; C, D)$ as a projection of one the path $A-B$ onto $C-D$; only edges shared between the paths will contribute non-zero amounts to this statistics. This projection analogy holds when we replace the populations by points in PCA-space, which enables us to study population models beyond discrete graphs. In this context, $F_4$ is simply an orthogonality test, an $F_4$ of zero means that the pairs of populations are orthogonal (i.e. independent) of another.

One difficulty in interpreting the results here is that they only hold when all PCs are considered. Yet, in my applications I found that two (for the World data set) and three PCs (for Western Eurasia) already provided a very good approximation for the majority of $F$-statistics, so that the results here should be applicable to at least continental-scale data sets.

>>>>

The two data sets I analyzed here suggest that two PCs (for the World data set) respectively three PCs (for the Western Eurasian data) already provide a very good approximation for $F_4$-statistics (Figures \ref{fig:weu}E, \ref{fig:world}E), reflecting the observation that frequently the first few PCs provide a good approximation of population structure. Thus, we may expect that whenever the first few PCs provide a good approximation of population structure, this results in well-approximated $F_4$-statistics. On the other hand, for admixture $F_3$-statistics, more PCs are needed (Figure \ref{fig:weu}BC), likely because these statistics do contain a term measuring local variation in the putative admixed population \citep{peter2016}, which is typically often incorporated on higher PCs. Thus, I would also expect individual-based analyses and $F_4$-statistics between closely related populations to require substantially more PCs as there local differentiation, which is typically reflected on higher PCs, becomes important.

@@

If sufficiently many PCs are considered, both $F_3$ and $F_4$-statistics do have simple interpretations in PCA-space. If used as a test for admixture, $F_3$ corresponds to testing whether the admixed population lies in a $n$-sphere between the potential source population in PCA-space. This result makes the informal notion that an admixed population should lie between its sources on a PCA-plot more precise. Furthermore, I show that while it is necessary for an admixed population to lie between its sources on a PCA plot, this is not a sufficient condition and unadmixed populations may also project inside the $n$-sphere. In contrast, if a population falls outside the $n$-sphere on any PCA-plot, this is sufficient to determine that this population has a positive $F_3$ statistic.

Interpretations of the outgroup-$F_3$-statistics and $F_4$-statistics in a PCA-framework rely heavily on the geometric concepts of projections and orthogonality: In a tree or admixture graph framework, we can loosely interpret $F_4(A, B; C, D)$ as the overlap of the path from $A$ to $B$ onto the path from $C$ to $D$, or equivalently as the  projection of $A-B$ onto $C-D$; only edges shared between the two paths will contribute non-zero amounts to this statistic. This interpretation holds when we replace the populations by points in PCA-space, and enables us to study population models beyond discrete graphs. Thus, the geometric framework enables us to expand the application of $F$-statistics beyond current uses, allows us to better understand the meaning of these statistics and helps us to avoid overinterpretations particularly in cases when populations structure is continuous.
